New cardinality estimation algorithms for HyperLogLog sketches

### 2016-12-27 paper/paper.tex
added outline to introduction section

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}. The first version of this paper was published on April 17, 2016.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle a second unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch. In particular, it is shown that it gives more accurate cardinality estimates for the union, intersection, or relative complements of two sets that are represented by HyperLogLog sketches compared to conventional techniques like the inclusion-exclusion principle. All the new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

%In contrast, HyperLogLog sketches (without historic inverse probability estimator) built from different sets of elements can be merged and the result is identical to a HyperLogLog sketch obtained by inserting all elements of the union set. Therefore, the HyperLogLog sketch inherently supports cardinality estimation of unions. However, there are also applications which require the cardinality estimation of intersections.

\section{HyperLogLog data structure}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision parameter that directly controls the relative error which scales like $1/\sqrt{\symNumReg}$. All registers start with zero initial value. Each element insertion potentially increases the value of one of these registers. The maximum value a register can reach is a natural bound given either by the output size of the used hash algorithm or the space that is reserved for a single register. Common implementations allocate up to 8 bits per register.

@@

\section{Maximum likelihood estimation}
We know from \cref{sec:depoissonization} that any unbiased estimator for the Poisson parameter is also an unbiased estimator for the cardinality. Moreover, we know that under suitable regularity conditions of the probability mass function the maximum likelihood estimator is asymptotically efficient \cite{Casella2002}. This means, if the number of registers $\symNumReg$ is large enough, the maximum likelihood method should give us an unbiased estimator for the cardinality.

@@

\section{Future work}
As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased estimator for the cardinality. We have shown that this approach works well for the improved raw estimator as well as for the maximum likelihood estimator even though both are only approximately unbiased. Therefore, it would be interesting what conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness, if used as cardinality estimator.

@@

\section{Conclusion}
We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation. For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation. The first uses the original estimator extended by theoretically motivated correction terms. The second is based on the maximum likelihood method and solves the corresponding optimization problem using the secant method. Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.

>>>>

\title{New cardinality estimation algorithms for HyperLogLog sketches}
\author{Otmar Ertl \\ otmar.ertl@gmail.com}
\begin{document}
\maketitle
\let\thefootnote\relax\footnotetext{This paper together with source code for all presented algorithms and simulations is available at \\ \url{https://github.com/oertl/hyperloglog-sketch-estimation-paper}. A first version of this paper was published on April 17, 2016.}
\begin{abstract}
This paper presents new methods to estimate the cardinalities of multisets recorded by HyperLogLog sketches. A theoretically motivated extension to the original estimator is presented that eliminates the bias for small and large cardinalities. Based on the maximum likelihood principle a second unbiased method is derived together with a robust and efficient numerical algorithm to calculate the estimate. The maximum likelihood approach can also be applied to more than a single HyperLogLog sketch. In particular, it is shown that it gives more accurate cardinality estimates for the union, intersection, or relative complements of two sets that are represented by HyperLogLog sketches compared to conventional techniques like the inclusion-exclusion principle. All the new methods are demonstrated and verified by extensive simulations.
\end{abstract}

@@

\subsection{Outline}
In \cref{sec:hyperloglog_data_structure} we give an introduction to HyperLogLog sketches and the corresponding update algorithm. We also discuss how the resulting state can be statistically described and approximated by a Poisson model.

In \cref{sec:cardinality_estimation} we describe the original cardinality estimation algorithm, its shortcomings at low and large cardinalities, and previous approaches to overcome them. We present a simple derivation of the original raw estimator and analyze the root cause for its limited operating range. We derive an improved version of the raw estimator, which leads to a new fast cardinality estimation algorithm that works over the full cardinality range as demonstrated by extensive simulations.

In \cref{sec:max_likelihood_estimation} we derive a second even more precise cardinality estimation algorithm based on the maximum likelihood principle which is again verified by simulations. As presented in \cref{sec:cardinality_estimation_set_intersections} the same approach can be generalized to two HyperLogLog sketches, which allows result cardinality estimation of set operations like intersections or complements. The simulation results show that the estimates are significantly better than those obtained by the inclusion-exclusion principle.

Finally, in \cref{sec:future_work} we discuss open problems and ideas for future work including HyperLogLog's potential use as locality-sensitve hashing algorithm before we conclude in \cref{sec:conclusion}.

\section{HyperLogLog data structure}
\label{sec:hyperloglog_data_structure}
The HyperLogLog algorithm uses a sketching data structure that consists of $\symNumReg$ registers. For performance reasons the number of registers $\symNumReg$ is chosen to be a power of 2, $\symNumReg = 2^\symPrecision$. $\symPrecision$ is the precision parameter that directly controls the relative error which scales like $1/\sqrt{\symNumReg}$. All registers start with zero initial value. Each element insertion potentially increases the value of one of these registers. The maximum value a register can reach is a natural bound given either by the output size of the used hash algorithm or the space that is reserved for a single register. Common implementations allocate up to 8 bits per register.

@@

\section{Maximum likelihood estimation}
\label{sec:max_likelihood_estimation}
We know from \cref{sec:depoissonization} that any unbiased estimator for the Poisson parameter is also an unbiased estimator for the cardinality. Moreover, we know that under suitable regularity conditions of the probability mass function the maximum likelihood estimator is asymptotically efficient \cite{Casella2002}. This means, if the number of registers $\symNumReg$ is large enough, the maximum likelihood method should give us an unbiased estimator for the cardinality.

@@

\section{Future work}
\label{sec:future_work}
As described in \cref{sec:depoissonization} an unbiased estimator for the rate in the Poisson model, is also unbiased estimator for the cardinality. We have shown that this approach works well for the improved raw estimator as well as for the maximum likelihood estimator even though both are only approximately unbiased. Therefore, it would be interesting what conditions on an approximately unbiased Poisson rate estimator are sufficient to guarantee approximate unbiasedness, if used as cardinality estimator.

@@

\section{Conclusion}
\label{sec:conclusion}
We have presented new algorithms for the estimation of cardinalities from HyperLogLog sketches based on the Poisson approximation. For the estimation from a single sketch, we have developed two fast algorithms that make use of the Poisson approximation. The first uses the original estimator extended by theoretically motivated correction terms. The second is based on the maximum likelihood method and solves the corresponding optimization problem using the secant method. Both algorithms are inherently unbiased and can be implemented without dependence on empirically determined bias correction data.
